Bridging the Gap Between Visual and Analytical Machine Learning Testing

### 2023-10-12 report/icst24.tex
feat(report): 2023-10-12 init implications section

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{array}

@@

\begin{abstract}
Testing ML systems is a highly interactive process which demands a
human-in-the-loop approach. In addition to writing tests for the code
base, practitioners are required to analyse and interpret several
visualisations using their domain expertise to validate if an ML
system satisfies the required set of functional and non-functional
properties. Visualisations are frequently used to qualitatively assess
various parts of an ML pipeline. However, implicit knowledge gained
from visualisations must be translated to explicit analytical tests
that fail when there is change in any component of the ML pipeline. We
conduct an empirical analysis of Jupyter notebooks to catalogue the
state-of-the-art mappings between ML visualisations and assertions. We
mine Github to collect Jupyter Notebooks that contain assertions
written in Python. We develop a novel methodology to identify 1764
notebooks which contain an assertion below a visualisation. We
manually analyse the 1.7K notebooks and identify 551 notebooks where
at least one pair of visualisation and assertion are semantically
related to one another. We further read and understand 90 notebooks in
depth to identify 26 qualitative and their corresponding quantitative
testing patterns. We showcase five such patterns in this paper and
plan to make the entire catalogue publicly available in the near
future.
\end{abstract}

@@

Visualisations are used throughout the ML development lifecycle to
test various properties of the ML system. In the early stages of the
ML development lifecycle, visualisations are extensively used for make
sense of the data and verify its statistical properties. During the
model development phase, visualisations are used to summarise metrics
and contrast different learning algorithms and iteratively fine-tune
ML models. Once the model is deployed in production, visualisations
are used to continually monitor their performance and trigger a new
training cycle once their performance drops below a certain threshold.

@@

In contrast to prior work done in ML testing, we approach this
pressing topic from a novel direction. We conduct an empirical
analysis of computational notebooks obtained from Github, to
understand the process of testing ML systems in practice. In
particular, we focus on the combination of a qualitative form of
testing (using visualisations) and a qualitative form of testing
(using analytical assertions). The research questions along with the
contributions of this paper are as follows.

@@

From the 551 notebooks, we further perform an in-depth analysis of 90
    notebooks. We identify 26 unique visualisation and assertion
    pairs. We discuss five such pairs in this paper and plan to
    release the full catalogue publicly in the near future.
\end{description}

@@

% TODO summarise literature on fairness testing
Bulk of the contributions on testing non-functional properties of ML
systems, focuses on fairness.

@@

\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}
{
  "metadata": {
    "kernel_info": {...},
    "language_info": {...},
  },
  "nbformat": 4,
  "nbformat_minor": 0,
  "cells": [ # dictionary of all cells in notebook ]
}
  \end{lstlisting}
  \captionof{lstlisting}{Top-level JSON format of Jupyter notebooks.
    See Listing~\ref{lst:cell-md},~\ref{lst:cell-code}
    and~\ref{lst:cell-output} for internal structure of various
    cells.}\label{lst:nbformat}
\end{minipage}
\hfill
\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}
{
  "cell_type": "markdown",
  "metadata": {},
  "source": [
    "# Hello world example\n",
    "The following cell contains python code for greeting people!\n"
  ]
}
  \end{lstlisting}
  \captionof{lstlisting}{JSON structure of markdown cells.}
  \label{lst:cell-md}
\end{minipage}
\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}
{
  "cell_type": "code",
  "metadata": {},
  "source": [
    "def greet(name):\n",
    "    print(f'Hello, {name}!')\n"
  ]
}
  \end{lstlisting}
  \captionof{lstlisting}{JSON structure of code cells.}
  \label{lst:cell-code}
\end{minipage}
\hfill
\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}
{
  "cell_type": "code",
  "metadata": {},
  "source": [...],
  "outputs": [
    {
      "output_type": "display_data",
      "data": {
        "image/png": "<base64-encoded-png-data>"
      }
    }
  ]
}
  \end{lstlisting}
  \captionof{lstlisting}{JSON structure of code cells with an image output.}
  \label{lst:cell-output}
\end{minipage}

@@

Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and
~\ref{lst:cell-output} present three examples of cells. Each cell is
represented as a dictionary with the \texttt{cell\_type},
\texttt{metadata} and \texttt{source} keys. Cells can be of three
types: ``markdown'', ``code'' or ``raw''. The contents of each cell
can be found within the \texttt{source} key, as a list of strings.

Outputs generated by a code cell are stored as a list within the
\texttt{outputs} key as seen in Listing~\ref{lst:cell-output}. Static
visualisations are stored as base64 encoded png images under the
\texttt{"image/png"} key, which is nested within the \texttt{data} key
inside \texttt{outputs}.

@@

\section{Methodology}\label{sec:method}
Figure~\ref{fig:method} presents an overview of all relevant steps
used in this study. The empirical analysis comprises of three distinct
phases namely: \textit{the data collection phase},
\textit{identification of related visualisation and assertion pairs}
and finally a \textit{complete analysis of the notebooks} to identify
the final set of 26 visualisation and assertion pairs.

@@

We mined public repositories from Github to collect Jupyter notebooks
that contain the keyword ``assert'' in them. We intentionally keep the
search criteria on Github broad by only checking for the appropriate
filetype and the presence of the ``assert'' keyword. This is because,
the ``assert'' keyword is used in python to define
tests. Additionally, popular python libraries which provide a testing
interface (such as nose, unittest and numpy), contain the same keyword
in their API. Our search criteria captures a large initial sample of
$54,070$ notebooks with a variety of testing statements. Our approach
also prevents the need to craft a custom search query based on an
exhaustive list of keywords that appear in all python testing
libraries.

We use the Github advanced search syntax~\footnote{REFME} to isolate
Jupyter Notebook. This translates to the following Github search
query: \texttt{filename:"Jupyter Notebooks"}. Additionally, we include
the ``assert'' keyword in the search query with an \texttt{AND}
operator. The final search query is as follows:
\texttt{filename:"Jupyter Notebooks" AND "assert"}.

The above search query results in approximately 54K notebooks. To
further reduce the sample size, we exclude notebooks that do not use
popular ML libraries. We perform a string search to identify notebooks
that import at least one of the ML libraries. The string patterns are
derived from the python module name used in the import statements. The
regex search query is as follows:
\texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.

We programmatically parse the JSON structure of the notebooks and
represent each notebook as a pandas dataframe for further analysis. We
map each cell as a row in the dataframe. For each cell, we store the
\texttt{cell\_type}, the \texttt{source} of the cell and the base64
encoded PNG string, into the corresponding columns of the
dataframe. We exclude notebooks that do not contain any visualisations
by checking if the image column of the dataframe is empty. The
internal structure of Jupyter notebooks is presented in
Section~\ref{sec:nbformat}.

@@

For RQ1, we are interested in finding empirical evidence of
visualisations and corresponding assertions used in the wild. To
answer this question, we further reduce our notebook sample size by
assuming a specific order in which the visualisation and assertion
code cells appear in the notebooks. Figure~\ref{fig:cell-arrangement}
presents a visual representation of various arrangements of code cells
that may occur in notebooks. Note that the visual representation omits
markdown cells since we only consider the order of code cells in our
analysis. In the actual notebook however, one or more markdown cells
may be present between the visualisation and assertion cells.

@@

We observed the best results in notebooks where the assertion cell was
not longer than 10 lines and was defined below the visualisation cell
(Arrangement 3.2). We observed that most notebook authors followed the
convention of writing their motivation for the visualisation in
a markdown cell, before writing the code for the visualisation. The
visualisations were often followed by another markdown cell to record
the author's observations. And finally, the observations were
translated to a analytical test using \texttt{``assert''} or other
testing methods from external libraries. The restriction on the number
of lines in the assertion cell was introduced to exclude instances
where the assertion cell contained a new class definition or several
helper functions that were not related to the visualisation above.

@@

The initial sample size of 54K notebooks from Github, was reduced to
the final sample size of 1.7K after applying the filters presented in
Section~\ref{sec:data-collect}. We analysed all 1.7K notebooks
manually to further identify 551 notebooks that contained at least one
pair of visualisation and assertions that were related to each
other. Table~\ref{tab:exclusion-criteria} presents the exclusion
criteria used in this phase of the analysis, along with the number of
notebooks that were excluded based on each criteria.

We manually analysed the source code of the visualisation and
assertion. And excluded visualisation and assertion pairs that were
not related to each other. We encountered several instances where the
tracing was not straightforward due to poor coding practices. In such
cases, we used one-shot prompt engineering~\cite{CITME} with the
ChatGPT 4.0 model to determine if the visualisation and assertion were
related to each other.

We excluded notebooks that did not train an ML model. We found several
notebooks authored on topics such as linear algebra, optimisation and
loss functions. Although these topics are related to ML, the
visualisation and assertion pairs found in these notebooks could not
be readily applied to a typically ML development workflow. Analysing
such notebooks also required a working knowledge of all mathematical
concepts related to ML. This was deemed beyond the scope of software
engineering research.

Finally, we excluded notebooks that were not authored in English or
contained faulty code that did not produce
a visualisation. Table~\ref{tab:exclusion-criteria} lists all
exclusion criteria used along with the number of notebooks that were
excluded from this study.

@@

\section{Results}\label{sec:result}

% TODO move the stuff for RQ1 from methodology here!

% NOTE the 5 pairs is to illustrate the 26 total pairs/approach we found for testing ML systems

% NOTE visual+analytical test is a new "strategy" for testing AI

This study identified 26 patterns of testing using visualisation and
assertion. This section highlights five interesting patterns in more
detail. We plan to release the entire catalogue of patterns publicly
in the near future.

\subsection{Decision boundary}\label{sec:svm}

% NOTE something about how the visualisation helps with interpretability of how the model is classifying the data?

% NOTE the author is using the visualisation to iteratively check/test how different kernels perform on the data; can also use it for hyper-parameter tuning?

% NOTE: visualisation can provide qualitative assessment of under/overfitting

% NOTE: better assertion: perhaps on the support vectors?

% NOTE assertion created was a single metric (performance): we need tools/guidelines to help practitioners capture what they see in visualisations; community failing to help practitioners what they are visualising

% NOTE assertion is not capturing the whole thing, only assert == we lose a lot of info, we need both

\begin{figure}
  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-04-linear.png}}
  \hfill
  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-04-rbf.png}}
  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}
  \label{fig:svm}
\end{figure}
\begin{lstlisting}[language=Python]
assert accuracy_score(y_test, pred) > 0.95
\end{lstlisting}
\captionof{lstlisting}{Assertion on the accuracy of the ML model.}\label{lst:svm}

@@

% NOTE maybe we need to ellaborate; the linear kernel was above the rbf; based on observations from the linear kernel, the author decides to try the rbf kernel; then after experimenting with hyper-parameters the author puts an assertion that captures the model's performance at that specific point in time.

@@

\subsection{Input Image, \texttt{predict\_proba}}

\begin{figure}
  \subcaptionbox{Random input image of a football from the test set.\label{fig:football}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-35a.png}}
  \hfill
  \subcaptionbox{Random input image of tennis balls from the test set.\label{fig:tennisball}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-35b.png}}
  \caption{Visualisation of random input images for a pre-trained image classifier.}
  \label{fig:balls}
\end{figure}

\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}[language=Python]
np.testing.assert_almost_equal(pred_proba, 2.0355723e-05, decimal=3)
  \end{lstlisting}
  \captionof{lstlisting}{Assertion below Figure~\ref{fig:football}. It checks if the probability estimate of the model is close to the specified value. The input image is of a football. Since the model is trained to detect tennis balls, the specified value is very small.}
  \label{lst:football}
\end{minipage}
\hfill
\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}[language=Python]
np.testing.assert_almost_equal(pred_proba, 0.9988895, decimal=3)
  \end{lstlisting}
  \captionof{lstlisting}{Assertion below Figure~\ref{fig:tennisball}. Compared to Listing~\ref{lst:football}, the specified value is much higher, since the input image contains tennis balls.}
  \label{lst:tennisball}
\end{minipage}

@@

\subsection{Normality}\label{sec:normal}

\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}[language=Python]
for feature in range(data_transformed.shape[1]):
  assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
  \end{lstlisting}
  \captionof{lstlisting}{Assertion to check that each feature in a dataset is normal. The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}
  \label{lst:kstest}
\end{minipage}
\hfill
\begin{minipage}{0.45\textwidth}
  \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
  \captionof{figure}{Visualisation of the distribution of a feature from the dataset.}
  \label{fig:kstest}
\end{minipage}

% NOTE assertion and viz are perfect fit;

% NOTE despite the analytical test, practitioners need to visualise this, we need visual confirmation; we cannot get rid of visualisations. we need both!

@@

% NOTE I want to say that we don't even think about the fact that we always use a histogram to do this! its second nature, we are taught to do this, without thinking why!

@@

\subsection{Coefficient of Determination}\label{sec:r2}

% NOTE this vis,assert is model agnostic!

% NOTE assertion captured what they visualised

% NOTE visualisation helps to check for outliers: we are humans, we look at visual patterns, quick and easier to observe fro visualisations

\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}[language=Python]
r2_gru = r2_score(y_test, y_pred)
assert r2_gru > 0.6
  \end{lstlisting}
  \captionof{lstlisting}{Assertion to check that the Coefficient of Determination ($R^2$) is higer than the specified threshold.}
  \label{lst:r2}
\end{minipage}
\hfill
\begin{minipage}{0.45\textwidth}
  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
  \captionof{figure}{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}
  \label{fig:r2}
\end{minipage}

@@

% TODO more context on r2_score, it fits a linear regression model and checks the blah blah blah
Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}. The author uses \texttt{r2\_score} metric from the scikit-learn library to quantify the performance of the model. The \texttt{r2\_score} method fits a linear regression model on the true and predicted labels and computed the Coefficient of Determination ($R^2$) based on the residuals of the linear regression model. The assertion checks that the $R^2$ score of the model is always higher than $0.6$. Since the best score possible is $1.0$, the specified threshold indicates that the author wants to ensure that $y_{pred}$ and $y_{true}$ are always linearly related to each other.

@@

\subsection{Normalization}

% TODO check the notebook again for implementation of normalization

\begin{figure}
  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-65a.png}}
  \hfill
  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-65b.png}}
  \caption{Visualisation of data before and after various data pre-processing techniques.}
  \label{fig:data-pre-process}
\end{figure}

\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}[language=Python]
assert scaled_data[scaled_data.argmax()] <= 1
assert scaled_data[scaled_data.argmix()] >= 0
  \end{lstlisting}
  \captionof{lstlisting}{Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}
  \label{lst:scale}
\end{minipage}
\hfill
\begin{minipage}{0.45\textwidth}
  \begin{lstlisting}[language=Python]
assert normalized_data[normalized_data.argmax()] < 10
assert normalized_data[normalized_data.argmin()] > -7
  \end{lstlisting}
  \captionof{lstlisting}{Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}
  \label{lst:normal}
\end{minipage}

@@

\section{Discussion}\label{sec:discuss}

% NOTE comment on the over-whelming presence of magic thresholds in ML
% testing! where are authors getting these numbers from? link this to
% requirements engineering?

% NOTE comment on the anti-patterns in testing that we observe in the 5 patterns presented

% NOTE link the qualitative and quantitative testing to the ML lifecycle: how some tests will let practitioner know when something changes

% NOTE comment on evolution? we have to look at the entire lifecycle of a visualisation: notebooks are run iteratively?

\subsection{New form of testing}
% NOTE some form of commentary on the fact that is form of qualitative and quantitative testing is happening but not that popular yet!

% NOTE so we need more tools to aid practitioners to encourage this form of testing and also help them translate qualitative properties into quantitative tests!

\subsection{From Implicit Expectations to Explicit Assertions}
% NOTE regeneration of the model does not change this observation at this point in time

% NOTE discuss the pros and cons of having a qualitative assessment: they are dense with useful information; humans need visual patterns!

% NOTE then show that we also need qualitative assessment that can capture our expectations; this is important so that the model never violates that we understand from the visualisations

% NOTE now is the time to talk about how assertions act as documentation of the domain expertise and knowledge extracted from the visualisation from the original creator; this also scales as the system complexity and composition of the team/organisation evolves!

% NOTE what are the benefits using visualisations to assess the
% quality of ML systems? They are packed with a lot of
% information. And humans are cognitively tuned to identify patterns
% in visualisations.

Visualisations condense valuable information that is easy to understand for humans.

\subsection{Something about magic thresholds}\label{sec:magic-threshold}
% NOTE we observe that most analytical test translates to comparing a value against a threshold; this indicates the open challenges in ML testing; either we don't yet know how to translate analytical tests from visualisations or we cannot do this in the first place and we must use a hybrid approach?

% NOTE this is a better place to talk about the magic threshold phenomenon; link this to the magic number smell in software engineering, but in ML it is very prevalent and an alternative is not always clear

% NOTE and then motivate how we need more tools/techniques to aid practitioners to extract/translate all the info from visualisations into qualitative assertions!

\subsection{Humans need visual confirmation}\label{sec:visual}
% NOTE even though analytical tests are sufficient; humans still prefer visual confirmation

% NOTE when working with image dataset, we observe in our catalogue that practitioners prefers to also visually check the input image

% NOTE more detailed discussion on how a lot of information can be condensed into a single visualition and how humans are good at observing visual patterns

>>>>

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{array}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent\emph{#1}
\end{framed}}

@@

\begin{abstract}

Testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 551 notebooks where at least one pair of visualisation and assertion are semantically related to one another. We further read and understand 90 notebooks in depth to identify 26 qualitative and their corresponding quantitative testing patterns. We showcase five such patterns in this paper and plan to make the entire catalogue publicly available in the near future.

% TODO what were the primary findings of this study?

\end{abstract}

@@

Visualisations are used throughout the ML development lifecycle to test various properties of the ML system. In the early stages of the ML development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties. During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models. Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.

@@

In contrast to prior work done in ML testing, we approach this pressing topic from a novel direction. We conduct an empirical analysis of computational notebooks obtained from Github, to understand the process of testing ML systems in practice. In particular, we focus on the combination of a qualitative form of testing (using visualisations) and a quantitative form of testing (using analytical assertions). The research questions along with the contributions of this paper are as follows.

@@

From the 551 notebooks, we further perform an in-depth analysis of 90 notebooks. We identify 26 unique visualisation and assertion pairs. We discuss five such pairs in this paper and plan to release the full catalogue publicly in the near future.

\end{description}

@@

% TODO summarise literature on fairness testing
%
Bulk of the contributions on testing non-functional properties of ML systems, focuses on fairness.

@@

\begin{lstlisting}[caption={Top-level JSON format of Jupyter notebooks. See Listing~\ref{lst:cell-md},~\ref{lst:cell-code} and~\ref{lst:cell-output} for internal structure of various cells.}, label={lst:nbformat}]
{
  "metadata": {
    "kernel_info": {...},
    "language_info": {...},
  },
  "nbformat": 4,
  "nbformat_minor": 0,
  "cells": [ # dictionary of all cells in notebook ]
}
\end{lstlisting}
\begin{lstlisting}[caption={JSON structure of markdown cells.}, label={lst:cell-md}]
{
  "cell_type": "markdown",
  "metadata": {},
  "source": [
    "# Hello world example\n",
    "The following cell contains python code for greeting people!\n"
  ]
}
\end{lstlisting}

\begin{lstlisting}[caption={JSON structure of code cells.}, label={lst:cell-code}]
{
  "cell_type": "code",
  "metadata": {},
  "source": [
    "def greet(name):\n",
    "    print(f'Hello, {name}!')\n"
  ]
}
\end{lstlisting}

\begin{lstlisting}[caption={JSON structure of code cells with an image output.}, label={lst:cell-output}]
{
  "cell_type": "code",
  "metadata": {},
  "source": [...],
  "outputs": [
    {
      "output_type": "display_data",
      "data": {
        "image/png": "<base64-encoded-png-data>"
      }
    }
  ]
}
\end{lstlisting}

@@

Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and ~\ref{lst:cell-output} present three examples of cells. Each cell is represented as a dictionary with the \texttt{cell\_type}, \texttt{metadata} and \texttt{source} keys. Cells can be of three types: ``markdown'', ``code'' or ``raw''. The contents of each cell can be found within the \texttt{source} key, as a list of strings.
Outputs generated by a code cell are stored as a list within the
\texttt{outputs} key as seen in Listing~\ref{lst:cell-output}. Static
visualisations are stored as base64 encoded png images under the
\texttt{"image/png"} key, which is nested within the \texttt{data} key
inside \texttt{outputs}.

@@

\section{Methodology}\label{sec:method}

Figure~\ref{fig:method} presents an overview of all relevant steps used in this study. The empirical analysis comprises of three distinct phases namely: \textit{the data collection phase}, \textit{identification of related visualisation and assertion pairs} and finally a \textit{complete analysis of the notebooks} to identify the final set of 26 visualisation and assertion pairs.

@@

We mined public repositories from Github to collect $54,070$ Jupyter notebooks that contain the keyword ``assert'' in them. We use the Github advanced search syntax~\footnote{REFME} to isolate Jupyter Notebook. This translates to the following Github search query: \texttt{filename:"Jupyter Notebooks"}. Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator. The final search query is as follows: \texttt{filename:"Jupyter Notebooks" AND "assert"}.

To further reduce the sample size, we exclude notebooks that do not use popular ML libraries. We perform a string search to identify notebooks that import at least one of the ML libraries. The string patterns are derived from the python module name used in the import statements. The regex search query is as follows: \texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.

We programmatically parse the JSON structure of the notebooks (more details regarding the internal structure of Jupyter notebooks is presented in Section~\ref{sec:nbformat}) and represent each notebook as a pandas dataframe for further analysis. We map each cell as a row in the dataframe. For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe. We exclude notebooks that do not contain any visualisations by checking if the image column of the dataframe is empty.

@@

To identify visualisation and assertion pairs that are semantically related to one another, we only consider notebooks in which the visualisation and assertion code cells appear in a specific order. Figure~\ref{fig:cell-arrangement} presents a visual representation of various arrangements of code cells that may occur in notebooks. Note that the visual representation omits markdown cells since we only consider the order of code cells in our analysis. In the actual notebook however, one or more markdown cells may be present between the visualisation and assertion cells.

@@

We observed the best results in notebooks where the assertion cell was not longer than 10 lines and was defined below the visualisation cell (Arrangement 3.2). We observed that most notebook authors followed the convention of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation. The visualisations were often followed by another markdown cell to record the author's observations. And finally, the observations were translated to a analytical test using \texttt{assert} or other testing methods from external libraries. The restriction on the number of lines in the assertion cell was introduced to exclude instances where the assertion cell contained a new class definition or several helper functions that were not related to the visualisation above.

@@

The initial sample size of 54K notebooks from Github, was reduced to the final sample size of 1.7K after applying the filters presented in Section~\ref{sec:data-collect}. We analysed all 1.7K notebooks manually to further identify 551 notebooks that contained at least one pair of visualisation and assertions that were related to each other.

We manually analysed the source code of the visualisation and assertion code cells. And excluded visualisation and assertion pairs that were not related to each other. We encountered several instances where the tracing was not straightforward due to poor coding practices. In such cases, we used one-shot prompt engineering~\cite{CITME} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.

We excluded notebooks that did not train an ML model. We found several notebooks authored on topics such as linear algebra, optimisation and loss functions. Although these topics are related to ML, the visualisation and assertion pairs found in these notebooks could not be readily applied to a typically ML development workflow. Analysing such notebooks also required a working knowledge of all mathematical concepts related to ML. This was deemed beyond the scope of this study.

Finally, we excluded notebooks that were not authored entirely in English or contained faulty code that did not produce a visualisation. Table~\ref{tab:exclusion-criteria} lists all exclusion criteria used along with the number of notebooks that were excluded from this study.

@@

\section{Results}\label{sec:result}
\subsection{RQ1. How frequently do ML practitioners formulate analytical tests from the visualisations they create to test ML systems?}\label{sec:result-rq1}

Of the 1764 Jupyter notebooks that were analysed in this study, we found 551 notebooks that contained at least one pair of visualisation and assertions which were semantically related to each other.

Our observations indicate that AI practitioners formulate analytical tests based on the information gained from the visualisations that they create to test ML systems. However, this is an emerging testing practice since only 31\% of the notebooks analysed in this study exhibit this testing practice.

\subsection{RQ2. What are the types of visualisation and assertion pairs typically used by ML practitioners when testing ML systems?}\label{sec:result-rq2}

This study identified 26 patterns of testing using visualisations and assertions. This section highlights five interesting patterns in more detail. We plan to release the entire catalogue of patterns publicly.

\subsubsection{Decision boundary}\label{sec:svm}

\begin{figure}
  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-linear.png}}
  \hfill
  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-rbf.png}}
  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}\label{fig:svm}
\end{figure}

\begin{lstlisting}[language=Python, caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]
assert accuracy_score(y_test, pred) > 0.95
\end{lstlisting}

@@

@@

\subsubsection{Input Image, \texttt{predict\_proba}}

\begin{figure}
  \subcaptionbox{Random input image of a football from the test set.\label{fig:football}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-35a.png}}
  \hfill
  \subcaptionbox{Random input image of tennis balls from the test set.\label{fig:tennisball}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-35b.png}}
  \caption{Visualisation of random input images for a pre-trained image classifier.}\label{fig:balls}
\end{figure}

\begin{lstlisting}[language=Python, caption={Assertion below Figure~\ref{fig:football}. It checks if the probability estimate of the model is close to the specified value. The input image is of a football. Since the model is trained to detect tennis balls, the specified value is very small.}, label={lst:football}]
np.testing.assert_almost_equal(pred_proba, 2.0355723e-05, decimal=3)
\end{lstlisting}

\begin{lstlisting}[language=Python, caption={Assertion below Figure~\ref{fig:tennisball}. Compared to Listing~\ref{lst:football}, the specified value is much higher, since the input image contains tennis balls.}, label={lst:tennisball}]
np.testing.assert_almost_equal(pred_proba, 0.9988895, decimal=3)
\end{lstlisting}

@@

\subsubsection{Normality}\label{sec:normal}

\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal. The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:ktest}]
for feature in range(data_transformed.shape[1]):
assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
\end{lstlisting}

\begin{figure}
  \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
  \caption{Visualisation of the distribution of a feature from the dataset.}\label{fig:kstest}
\end{figure}

@@

@@

\subsubsection{Coefficient of Determination}\label{sec:r2}

\begin{lstlisting}[language=Python, caption={Assertion to check that the Coefficient of Determination ($R^2$) is higer than the specified threshold.}, label={lst:r2}]
r2_gru = r2_score(y_test, y_pred)
assert r2_gru > 0.6
\end{lstlisting}

\begin{figure}
  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
  \caption{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
\end{figure}

@@

Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}. The author uses \texttt{r2\_score} metric from the scikit-learn library to quantify the performance of the model. The \texttt{r2\_score} method fits a linear regression model on the true and predicted labels and computed the Coefficient of Determination ($R^2$) based on the residuals of the linear regression model. The assertion checks that the $R^2$ score of the model is always higher than $0.6$. Since the best score possible is $1.0$, the specified threshold indicates that the author wants to ensure that $y_{pred}$ and $y_{true}$ are always linearly related to each other.

@@

\subsubsection{Normalization}

\begin{figure}
  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=\linewidth]{../catalogue/select-65a.png}}
  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=\linewidth]{../catalogue/select-65b.png}}
  \caption{Visualisation of data before and after various data pre-processing techniques.}
  \label{fig:data-pre-process}
\end{figure}

\begin{lstlisting}[language=Python, caption={Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}, label={lst:scale}]
assert scaled_data[scaled_data.argmax()] <= 1
assert scaled_data[scaled_data.argmix()] >= 0
\end{lstlisting}

\begin{lstlisting}[language=Python, caption={Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}, label={lst:normal}]
assert normalized_data[normalized_data.argmax()] < 10
assert normalized_data[normalized_data.argmin()] > -7
\end{lstlisting}

@@

\section{Implications}\label{sec:discuss}

This study mined 54K Jupyter Notebooks that contained an assertion. Further analysis identified 1.7K notebooks which contained an assertion directly below a visualisation. Of the 1.7K notebooks, 551 notebooks contained a visualisation and assertion pair which were semantically related to each other.A deeper analysis of 90 Jupyter notebooks revealved 26 unique ML testing tasks which utilised a combination of visualisations and assertions.

The results from RQ2 provide emprical evidence to support the use of visualisations to test various properties of ML systems.  Visualisations are an efficient and effective medium to present complex information, that is easier to interpret by humans with the aid of visual cues. Visualisations used to test ML system properties however, demand manual verification effort. Information gained from visualisations and subsequent decisions made based on this information, must be manually verified in every iteration of the ML development cycle.

% NOTE link to accumulation of technical debt due to highly experimental nature, and constant change in the system

Assertions or analytical tests extracted from ML visualisations can reduce this manual verification effort. Assertions encapsulate the knowledge gained from the visualisation with regards to the property of the ML system which is currently under test. Formal assertions document the observations made by the AI practitioner regarding the model or the data, at a specific point in time. These observations should not be violated by the future model development cycles unless the requirements of the ML system are changed. Assertions scale across organizational or personnel changes as future AI practitioners can refer to the assertions to understand how the visualisations were previously interpreted.

Results for RQ1 indicate that combining ML visualisations with analytical assertions is an emerging testing practice within the ML community. However in results from RQ2, we see several examples of assertions that fail to capture all the information gained from its visual counter-part. Our observations indicate a need to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations. We find the need for automated tools that make this translation process easier for AI practitioners.
